### 2023-02-08 paper-thesis/chapters/01-introduction.tex
Fix chapter references

Furthermore, the thesis contains some discussions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (Chapter \ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell}) and LambdaPi (Chapter \ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (Chapter \ref{ch:ergonomics})
	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:conclusion})
\end{itemize}

\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).

>>>>

Furthermore, the thesis contains some discussions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
\end{itemize}

\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (chapter~\ref{chap:bg-dp}) and the Calculus of Constructions (chapter~\ref{chap:bg-spoofax}).

### 2023-02-08 paper-thesis/chapters/03-background-dependent-types.tex
Fix chapter references

Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}. Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition. Thus if our type system is sound\footnote{The type system we will define in chapter \ref{chap:baselang} is not sound, we will fix this in chapter \ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.

@@

The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter \ref{chap:baselang}.

>>>>

Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}. Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition. Thus if our type system is sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not sound, we will fix this in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.

@@

The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter~\ref{chap:baselang}.

### 2023-02-08 paper-thesis/chapters/10-coc.tex
Fix chapter references

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}. We only use a single type of edge, called \verb|P| (parent) edges. We also only have a single relation, called \verb|name|. The relation associates a \verb|NameEntry| with each name in the scope graph. The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NType : Expr -> NameEntry
    NSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else. For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter \ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.

>>>>

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter~\ref{chap:bg-spoofax}. We only use a single type of edge, called \verb|P| (parent) edges. We also only have a single relation, called \verb|name|. The relation associates a \verb|NameEntry| with each name in the scope graph. The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NType : Expr -> NameEntry
    NSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else. For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter~\ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.

### 2023-02-08 paper-thesis/chapters/20-adding-booleans.tex
Improve booleans chapter
Work on CoC chapter
Fix chapter references

\begin{lstlisting}
BoolType : Expr
BoolFalse : Expr
BoolTrue : Expr
BoolIf : Expr * Expr * Expr -> Expr
\end{lstlisting}
Then, the rules for beta reducing the language are in figure \ref{fig:bool-rules-beta}. There is one particularly interesting case, that is how to beta-reduce if statements. Converting this rule to Statix is not entirely trivial, since you need to choose which rule to apply based on what \verb|c| evaluates to. A new rule is needed, which has 3 cases. One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution). These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter \ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)

\begin{lstlisting}
betaReduceHead((s, BoolIf(c, b1, b2)), t) = 
  betaReduceHeadIf(s, betaReduceHead((s, c), []), c, b1, b2, t).
betaReduceHeadIf(s, (_, BoolTrue()), _, b1, _, t) = 
  betaReduceHead((s, b1), t).
betaReduceHeadIf(s, (_, BoolFalse()), _, _, b2, t) = 
  betaReduceHead((s, b2), t).
betaReduceHeadIf(s, _, c, b1, b2, t) = 
  rebuild((s, BoolIf(c, b1, b2)), t).
\end{lstlisting}

@@

	\inferrule{
		\bhr
		{ \scope{s}{c} }
		{ [] }
		{ \scope{s'}{\co{BoolTrue}()} }
		\\\bhr
		{ \scope{s}{b1} }
		{ t }
		{ \scope{s''}{b1'} }
	} {
		\bhr
		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
		{ t }
		{ \scope{s}{b1'}}
	}

	\inferrule{
		\bhr
		{ \scope{s}{c} }
		{ [] }
		{ \scope{s'}{\co{BoolFalse}()} }
		\\\bhr
		{ \scope{s}{b2} }
		{ t }
		{ \scope{s''}{b2'} }
	} {
		\bhr
		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
		{ t }
		{ \scope{s}{b2'}}
	}
\end{mathpar}
\caption{Rules for beta head reducing booleans}
\label{fig:bool-rules-beta}
\end{figure}

@@

	\inferrule{
		\toe{s}{c}{ct}
		\\\beq{ct}{\co{BoolType}()}
		\\\\
		\\\toe{s}{b1}{tb1}
		\\\toe{s}{b2}{tb2}
		\\\beq{tb1}{tb2}
	} {
		\toe{s}{\co{BoolIf}(c, b1, b2)}{tb1}
	}
\end{mathpar}
\caption{Rules for type checking booleans}
\label{fig:bool-rules-typecheck}
\end{figure}

@@

\begin{figure}[ht]
	\begin{mathpar}
		\inferrule{
			\bhr
			{\scope{s}{b}}
			{a}
			{\scope{s'}{b'}}
		}{
			\bhr
			{\scope{s}{\co{Postulate}(n, t, b)}}
			{a}
			{\scope{s'}{b'}}
		}

		\inferrule{
			\toe{s}{t}{tt}
			\\\beq{tt}{\co{Type}()}
			\\\\\bred{s}{t}{t'}
			\\\toe{\co{sPutType}(s, n, t')}{b}{bt}
		} {
			\toe{s}{\co{Postulate}(n, t, b)}{bt}
		}
	\end{mathpar}
	\caption{Rules for postulates}
	\label{fig:postulate-rules}
\end{figure}

@@

\section{Extensibility of the approach}

@@

In the following chapters, we're going to be extending the language further with these more interesting features using this approach:
\begin{itemize}
	\item Inference (chapter \ref{ch:inference})
	\item Inductive Datatypes (chapter \ref{ch:datatypes})
	\item Universes (chapter \ref{ch:universes})
\end{itemize}

>>>>

\begin{lstlisting}
BoolType : Expr
BoolFalse : Expr
BoolTrue : Expr
BoolIf : Expr * Expr * Expr -> Expr
\end{lstlisting}
Then, the rules for beta reducing the language are in figure \ref{fig:bool-rules-beta}. There is one particularly interesting case, that is how to beta-reduce if statements. Converting this rule to Statix is not entirely trivial, since you need to choose which rule to apply based on what \verb|c| evaluates to. A new rule is needed, which has 3 cases. One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution). These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)

\begin{lstlisting}
betaReduceHead((s, BoolIf(c, b1, b2)), ps) = 
  betaReduceHeadIf(s, betaReduceHead((s, c), []), c, b1, b2, ps).
betaReduceHeadIf(s, (_, BoolTrue()), _, b1, _, ps) = 
  betaReduceHead((s, b1), ps).
betaReduceHeadIf(s, (_, BoolFalse()), _, _, b2, ps) = 
  betaReduceHead((s, b2), ps).
betaReduceHeadIf(s, _, c, b1, b2, ps) = 
  rebuild((s, BoolIf(c, b1, b2)), ps).
\end{lstlisting}

@@

	\\

	\inferrule{
		\bhr
		{ \scope{s}{c} }
		{ [] }
		{ \scope{s'}{\co{BoolTrue}()} }
		\\\\\bhr
		{ \scope{s}{b1} }
		{ \overline{p} }
		{ \scope{s''}{b1'} }
	} {
		\bhr
		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
		{ \overline{p} }
		{ \scope{s}{b1'}}
	}

	\inferrule{
		\bhr
		{ \scope{s}{c} }
		{ [] }
		{ \scope{s'}{\co{BoolFalse}()} }
		\\\\\bhr
		{ \scope{s}{b2} }
		{ \overline{p} }
		{ \scope{s''}{b2'} }
	} {
		\bhr
		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
		{ \overline{p} }
		{ \scope{s}{b2'}}
	}

	\inferrule{
	\bhr
	{ \scope{s}{c} }
	{ [] }
	{ \scope{s'}{c'} }
	\\\\ c' \neq \co{BoolTrue}()
	\\ c' \neq \co{BoolFalse}()
} {
	\bhr
	{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
	{ \overline{p} }
	{ \co{rebuild}((s, \co{BoolIf}(c, b1, b2)), \overline{p}) }
}
\end{mathpar}
\caption{Rules for beta head reducing booleans}
\label{fig:bool-rules-beta}
\end{figure}

@@

	\inferrule{
		\toe{s}{c}{t_c}
		\\\beq{t_c}{\co{BoolType}()}
		\\\\
		\\\toe{s}{b1}{t_{b1}}
		\\\toe{s}{b2}{t_{b2}}
		\\\beq{t_{b1}}{t_{b2}}
	} {
		\toe{s}{\co{BoolIf}(c, b1, b2)}{t_{b1}}
	}
\end{mathpar}
\caption{Rules for type checking booleans}
\label{fig:bool-rules-typecheck}
\end{figure}

@@

\begin{figure}[ht]
	\begin{mathpar}
		\inferrule{
			\bhr
			{\scope{s}{b}}
			{\overline{p}}
			{\scope{s'}{b'}}
		}{
			\bhr
			{\scope{s}{\co{Postulate}(n, t, b)}}
			{\overline{p}}
			{\scope{s'}{b'}}
		}

		\inferrule{
			\toe{s}{t}{t_t}
			\\\beq{t_t}{\co{Type}()}
			\\\\\bred{s}{t}{t'}
			\\\toe{\co{sPutType}(s, n, t')}{b}{t_b}
		} {
			\toe{s}{\co{Postulate}(n, t, b)}{t_b}
		}
	\end{mathpar}
	\caption{Rules for postulates}
	\label{fig:postulate-rules}
\end{figure}

@@

\section{Extensibility of the Approach}

@@

In the following chapters, we're going to be extending the language further with these more interesting features using this approach:
\begin{itemize}
	\item Inference (chapter~\ref{ch:inference})
	\item Inductive Datatypes (chapter~\ref{ch:datatypes})
	\item Universes (chapter~\ref{ch:universes})
\end{itemize}

### 2023-02-08 paper-thesis/chapters/30-comparing-with-haskell.tex
Add error annotations
Work on CoC chapter
Fix chapter references

In this chapter, we implement the language defined in chapter \ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix. We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.

@@

\section{Defining environments}

Next, we need to decide how we want to store the environment. We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter \ref{chap:baselang}. We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0. Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.
\begin{lstlisting}
data EnvEntry = NType Expr | NSubst SExpr
type Env = [EnvEntry]
type SExpr = (Env, Expr)
\end{lstlisting}

@@

\section{Defining beta reduction}

@@

The base implementations of the language are quite similar. One difference is the way the definitions are clustered\todo{not right word?}. In Statix, one could put each language construct in a separate file, keeping the definitions for that construct together. In the Haskell implementation this is not easily possible\footnote{We could still define a function that then calls the actual implementation in the separate files, but this is still inferior to the Statix implementation.}, since function definitions can not be split over a file.

Another advantage that Statix has is that it has first-order inference built-in, which makes implementing a basic form of inference as described in chapter \ref{ch:inference} way easier. However, if we want a more complex form of inference then an implementation in Haskell would be better, since it is a more expressive language and it has more libraries available.

Finally, creating the language in Spoofax automatically gives us editor services such as code highlighting and semantic autocompletion, as dicusssed in chapter \ref{chap:editor-services}. Implementing these in Haskell would require some work.

>>>>

In this chapter, we implement the language defined in chapter~\ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix. We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.

@@

\section{Defining Environments}

Next, we need to decide how we want to store the environment. We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter~\ref{chap:baselang}. We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0. Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.
\begin{lstlisting}
data EnvEntry = NType Expr | NSubst SExpr
type Env = [EnvEntry]
type SExpr = (Env, Expr)
\end{lstlisting}

@@

\section{Defining Beta Reduction}

@@

The base implementations of the language are quite similar. One difference is the way the definitions are spread. In Statix, one could put each language construct in a separate file, keeping the definitions for that construct together. In the Haskell implementation this is not easily possible\footnote{We could still define a function that then calls the actual implementation in the separate files, but this is still inferior to the Statix implementation.}, since function definitions can not be split over a file.

Another advantage that Statix has is that it has first-order inference built-in, which makes implementing a basic form of inference as described in chapter~\ref{ch:inference} way easier. However, if we want a more complex form of inference then an implementation in Haskell would be better, since it is a more expressive language and it has more libraries available.

Finally, creating the language in Spoofax automatically gives us editor services such as code highlighting and semantic autocompletion, as dicusssed in chapter~\ref{chap:editor-services}. Implementing these in Haskell would require some work.

### 2023-02-08 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Work on CoC chapter
Fix chapter references

In this chapter we will be implementing the language defined in chapter \ref{chap:baselang} in lambdapi, a proof assistant based on the $\lambda \Pi$-calculus modulo rewriting~\cite{BoespflugCH12}.

\section{Defining symbols}

@@

\section{Reduction rules}

@@

\section{\label{ch:comp-lambdapi:let}Defining let bindings}

>>>>

In this chapter we will be implementing the language defined in chapter~\ref{chap:baselang} in lambdapi, a proof assistant based on the $\lambda \Pi$-calculus modulo rewriting~\cite{BoespflugCH12}.

\section{Defining Symbols}

@@

\section{Reduction Rules}

@@

\section{\label{ch:comp-lambdapi:let}Defining Let Bindings}

### 2023-02-08 paper-thesis/chapters/40-discussing-statix.tex
Fix chapter references

\begin{enumerate}
	\item Spoofax is still slow, even though its performance has improved over the past years. On a high-end computer\footnote{Intel i7 6700k @ 4.2GHz, 16 GB Ram @ 2133MHz, Arch Linux @ Jan 2023, Spoofax 3 v0.19.2}, we took the following measurements.
	\begin{itemize}
		\item Full build (non-incremental) takes 55 seconds
		\item Incremental build changing only one Statix file takes 5 seconds
		\item Running all 171 tests takes 18 seconds
	\end{itemize}
	While these numbers are not unworkably high, they are still magnitudes higher than the implementation in Haskell (chapter \ref{ch:comp-haskell}) and other general-purpose languages. Improving these numbers would make the experience better.

>>>>

\begin{enumerate}
	\item Spoofax is still slow, even though its performance has improved over the past years. On a high-end computer\footnote{Intel i7 6700k @ 4.2GHz, 16 GB Ram @ 2133MHz, Arch Linux @ Jan 2023, Spoofax 3 v0.19.2}, we took the following measurements.
	\begin{itemize}
		\item Full build (non-incremental) takes 55 seconds
		\item Incremental build changing only one Statix file takes 5 seconds
		\item Running all 171 tests takes 18 seconds
	\end{itemize}
	While these numbers are not unworkably high, they are still magnitudes higher than the implementation in Haskell (chapter~\ref{ch:comp-haskell}) and other general-purpose languages. Improving these numbers would make the experience better.

### 2023-02-08 paper-thesis/chapters/99-conclusion.tex
Improve the introduction
Fix chapter references

\chapter{Conclusion}

@@

\paragraph{RQ1: Can the Calculus of Constructions be implemented in Statix?}
As we demonstrated in chapter \ref{chap:baselang}, the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph. We defined the beta-reduction, beta-equality and type checking rules and then converted them to Statix code. Beta reduction was defined using a Krivine machine. In chapter \ref{chap:namecolls} we solved the variable capture problem which the naive implementation from chapter \ref{chap:baselang} suffered from, by using scoped names.

\paragraph{RQ2: Is the implementation is easily extendable?}
We showed in chapter \ref{chap:bools} that the implementation is easily extendable, by extending it with booleans, postulate, and type assertions. We define a four step process that can be used to extend the language, which we also use to answer the remaining research questions.

\paragraph{RQ3: Can we add inference to the implementation?}
In chapter \ref{ch:inference} we discuss how to add inference to the implementation. If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference. We implemented this algorithm in a concise way.

\paragraph{RQ4: Can we add support for inductive data types to the implementation?}
Chapter \ref{ch:datatypes} adds support for inductive data types with parameters and indices. We give steps for type-checking data type declarations. Additionally, we give each data type an eliminator and show how to type-check and beta reduce eliminators. Finally, we show that positivity checking can be implemented concisely.

\paragraph{RQ5: Can we add support for universes to the implementation?}
In chapter \ref{ch:universes} we showed that we can add universes to the language. This can be done easily and concisely.

>>>>

\chapter{Conclusion}
\label{ch:conclusion}

@@

\paragraph{RQ1: Can the Calculus of Constructions be implemented in Statix?}
As we demonstrated in chapter~\ref{chap:baselang}, the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph. We defined the beta-reduction, beta-equality and type checking rules and then converted them to Statix code. Beta reduction was defined using a Krivine machine. In chapter~\ref{chap:namecolls} we solved the variable capture problem which the naive implementation from chapter~\ref{chap:baselang} suffered from, by using scoped names.

\paragraph{RQ2: Is the implementation is easily extendable?}
We showed in chapter~\ref{chap:bools} that the implementation is easily extendable, by extending it with booleans, postulate, and type assertions. We define a four step process that can be used to extend the language, which we also use to answer the remaining research questions.

\paragraph{RQ3: Can we add inference to the implementation?}
In chapter~\ref{ch:inference} we discuss how to add inference to the implementation. If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference. We implemented this algorithm in a concise way.

\paragraph{RQ4: Can we add support for inductive data types to the implementation?}
chapter~\ref{ch:datatypes} adds support for inductive data types with parameters and indices. We give steps for type-checking data type declarations. Additionally, we give each data type an eliminator and show how to type-check and beta reduce eliminators. Finally, we show that positivity checking can be implemented concisely.

\paragraph{RQ5: Can we add support for universes to the implementation?}
In chapter~\ref{ch:universes} we showed that we can add universes to the language. This can be done easily and concisely.
